{\fontfamily{ptm}\selectfont Nesterov Type Momentum Methods}

### 2024-07-04 Notebook LaTEX/Momentum Accelerations/main.tex
>w< good stuff are coming

\section{Accelerated gradient descent and PPM}
    Recent works from Ahn and Sra \cite{ahn_understanding_2022} and Nesterov \cite{nesterov_lectures_2018} inspired content in this section.
    Ahn explored the interpretation of Nesterov acceleration via PPM. 
    They proposed the idea of ``similar triangle" for unifying all varieties of Nesterov accelerated gradient. 
    They used PPM to derive several variations of the Nesterov accelerated gradient algorithms. 
    Finally, they refurnished \hyperref[thm:lower_approx_ppm_convergence]{theorem \ref*{thm:lower_approx_ppm_convergence}} for the proof of convergence rate for the accelerated gradient.
    Their analysis results in relatively simple arguments that exhibits powerful extensions to several variants of the Nesterov accelerated gradient. 
    \par\noindent
    Interestingly, the Nesterov accelerated gradient applies to PPM too; Guler \cite{guler_new_1992} did it two decades ago. 
    He uses the idea of a Nesterov acceleration sequence faithfully. 
    One recent development of the accelerated PPM is an algorithmic framework named: ``Universal Catalyst acceleration", proposed by Lin et al \cite{lin_universal_2015}. 
    It is an application of Guler's work in the context of variance-reduction stochastic gradient algorithms for machine learning.

    \subsection{Varieties of Nesterov accelerated gradient}
        \subsubsection{AG Abstract Forms}
            In this section, we list different varieties of the Nesterov accelerated method. 
            \begin{definition}[Nestrov 2.2.7]\label{def:Nes2.2.7}
                Let $f$ be a $L$ Lipschitz smooth and $\mu\ge 0$ strongly convex function. 
                Choose $x_0$, $\gamma_0 > 0$, set $v_0 = x_0$, for iteration $k\ge 0$, it
                \begin{enumerate}
                    \item[1.] computes $\alpha_k \in (0, 1)$ by solving $L\alpha_k^2 = (1 - \alpha_k)\gamma_k + \alpha_k \mu$; 
                    \item[2.] sets $\gamma_{k + 1} = (1 - \alpha_k)\gamma_k + \alpha_k \mu$;
                    \item[3.] chooses $y_k = (\gamma_k + \alpha_k \mu)(\alpha_k \gamma_k v_k + \gamma_{k + 1}x_k)$. Compute $f(y_k)$ and $\nabla f(y_k)$; 
                    \item[4.] finds $x_{k + 1}$ such that $f(x_{k + 1}) \le f(y_k) - (2L)^{-1} \Vert \nabla f(y_k)\Vert^2$; 
                    \item[5.] sets $v_{k + 1} = \gamma_{k+1}^{-1}((1 - \alpha_k)\gamma_kv_k + \alpha_k \mu y_k - \alpha_k \nabla f(y_k))$. 
                \end{enumerate}
            \end{definition}
            \begin{remark}
                This is in Nesterov's book \cite[(2.2.7)]{nesterov_lectures_2018}. 
                For more context, the sequence $\alpha_k \in (0, 1)$ such that $\sum_{k = 1}^{\infty}\alpha_k = \infty$, and recursively we have $\lambda_{k + 1} = (1 - \alpha_k)\lambda_k$. 
                the sequence $\lambda_k$ is called an Nesterov estimating sequence. 
                It is the most generic algorithm in his book about accelerated gradient method. 
                The genericity of the algorithm is provided by item 4., which is the a special case of the smooth descent lemma.

@@

@@

        \subsubsection*{Useful Observations about these Generic forms}

@@

                $L$-Lipschitz continuous, 

@@

                Similarly, when $\mu = 0$, we have from the first step that $v_{k + 1} = v_k - \alpha_k \nabla f(y_k)$, which is the same as the AG generic PPM form where 

@@

            performing PPM on $\phi_t$ produces the PPM Lyapunov inequality from 
            \hyperref[thm:ppm_descent_ineq]{theorem \ref*{thm:ppm_descent_ineq}},

@@

                =: \Upsilon_{1, t + 1}^{\text{AG}}

@@

                - \frac{1}{2}\Vert x_{t} - x_*\Vert^2 

@@

        \begin{theorem}[Generic AG Convergence]\label{thm:generic_smooth_ag_convergence}
            If there is a choice of $\eta_i, \tilde \eta_i$ in \hyperref[def:agg_ppm]{definition \ref*{def:agg_ppm}} such that we have 
            \begin{align*}
                \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)\Upsilon_{2, t + 1}^{\text{AG}} + \Upsilon_{2, t + 1}^{\text{AG}} \le 0, 
            \end{align*}
            then with 
            \begin{align*}
                \Phi_t &= \left(
                    \sum_{i = 1}^{t} \tilde\eta_{t + 1}
                \right) (f(z_t) - f(x_*)) + \frac{1}{2}\Vert x_t - x_*\Vert^2 \quad \forall t \in \NN
                \\
                \Phi_0 &= \frac{1}{2}\Vert x_t - x_*\Vert^2, 
            \end{align*}
            we have $\Phi_{t + 1} - \Phi_t \le 0$, which allows for a convergence rate of $\mathcal O \left(\sum_{i = 1}^{T} \eta_i^{-1}\right)$ for $f(z_T) - f(x_*)$. 
        \end{theorem}
        \begin{proof}
            By definition we have
            {\footnotesize
            \begin{align*}
                \Phi_{t + 1} - \Phi_t 
                &= 
                \left(
                    \sum_{i = 1}^{t+1} \tilde\eta_{i}
                \right) (f(z_{t + 1}) - f(x_*)) 
                - 
                \left(
                    \sum_{i = 1}^{t} \tilde\eta_{i}
                \right) (f(z_{t}) - f(x_*)) 
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                - \frac{1}{2}\Vert x_{t + 1} - x_*\Vert^2
                \\
                &= 
                \tilde \eta_{t + 1} (f(z_{t + 1}) - f(z_*))
                +
                \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)(f(z_{t + 1}) - f(z_t))
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                - \frac{1}{2}\Vert x_{t + 1} - x_*\Vert^2
                \\
                &= \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)\Upsilon_{2, t + 1}^{\text{AG}} + \Upsilon_{2, t + 1}^{\text{AG}} \le 0. 
            \end{align*}
            }
            The derivation for the convergence raet is direct and similar to 
            \hyperref[thm:ppm_convergence_rate]{theorem \ref*{thm:ppm_convergence_rate}} 
            Therefore, if we can identify parameter for the generic algorithm that asserts the condition above, then we have convergence for the algorithm. 
        \end{proof}

    \subsection{Specific Varieties of Accelerated Gradient Method}
        In this section we state some specific variant of the Nesterov accelerated gradient method.

>>>>

\section{Accelerated gradient descent and PPM}
    Recent works from Ahn and Sra \cite{ahn_understanding_2022} and Nesterov \cite{nesterov_lectures_2018} inspired content in this section.
    Ahn explored the interpretation of Nesterov acceleration via PPM. 
    They proposed the idea of ``similar triangle" for unifying all varieties of Nesterov accelerated gradient. 
    They used PPM to derive several variations of the Nesterov accelerated gradient algorithms. 
    Finally, they refurnished \hyperref[thm:lower_approx_ppm_convergence]{theorem \ref*{thm:lower_approx_ppm_convergence}} for the proof of convergence rate for the accelerated gradient.
    Their analysis results in relatively simple arguments that exhibits powerful extensions to several variants of the Nesterov accelerated gradient.

    \par\noindent
    Interestingly, the Nesterov accelerated gradient applies to PPM too; Guler \cite{guler_new_1992} did it two decades ago. 
    He uses the idea of a Nesterov acceleration sequence faithfully. 
    One recent development of the accelerated PPM is an algorithmic framework named: ``Universal Catalyst acceleration", proposed by Lin et al \cite{lin_universal_2015}. 
    It is an application of Guler's work in the context of variance-reduction stochastic gradient algorithms for machine learning.

    \par\noindent
    \textcolor{red}{We state our contributions}. 
    In \hyperref[sec:AG_varieties]{section \ref*{sec:AG_varieties}} 
    We stated different forms of accelerated gradient that appeared in the literatures and the equivalences between their forms. 
    They are abstract because the choice of stepsize parameters in the algorithms are unspecified. 
    In \hyperref[sec:AG_useful_information]{section \ref*{sec:AG_useful_information}}
    we prove and interpret some of the unproved claims made by Ahn, Sra \cite{ahn_understanding_2022} as good exercises and for the peace of the mind.
    \hyperref[lemma:smooth_agg_lyapunov_upper_bound]{Lemma \ref*{lemma:smooth_agg_lyapunov_upper_bound}} 
    rephrase content in the appendix in work by Ahn, Sra\cite{ahn_understanding_2022}. 
    It is our attempt at rephrasing the proofs to create a layer of abstraction so it can be reused for multiple proofs for several variants.

    \subsection{Preliminaries}
        We introduce some additional lemma that are crucial to the derivations of non-smooth accelerated gradient method. 
        \begin{definition}[The Gradient Mapping]
            \label{def:gradient_mapping}
            Let $g = f + g$ where $f$ is $L$-Lipschitz smooth and convex, $g$ is convex. 
            Define the proximal gradient operator
            $$
                \mathcal T_L(x) := \hprox_{L^{-1}g}(x - L^{-1}\nabla f(x)),
            $$
            then the gradient mapping is defined as
            $$
                \mathcal G_L(x) = L(x - \mathcal T_L(x)). 
            $$
        \end{definition}
        \begin{remark}
            The name "gradient mapping" comes from \cite[(2.2.54)]{nesterov_lectures_2018}, however, Nesterov was referring to only the case when $g$ is an indicator function of a convex set in his writing. 
        \end{remark}

        \begin{lemma}[Gradient Mapping Approximates Subgradient]
            \label{lemma:grad_map_lemma_first}
            Continue from 
            \hyperref[def:gradient_mapping]{definition \ref*{def:gradient_mapping}}, 
            the gradient mapping satisfies
            \begin{align*}
                x^+ &= \mathcal T_L(x), 
                \\
                L(x - x^+) &\in  \nabla f(x) + \partial g(x^+) \ni \mathcal G_L(x). 
            \end{align*}
        \end{lemma}
        \begin{proof}
            \begin{align*}
                x^+ &= [I + L^{-1}\partial g]^{-1}\circ [I - L^{-1}\nabla f](x)
                \\
                [I + L^{-1}\partial g](x^+) &\ni [I - L^{-1}\nabla f](x)
                \\
                x^+ + L^{-1}\partial g(x^+) &\ni x - L^{-1}\nabla f(x)
                \\
                x^+ - x + L^{-1}\partial g(x^+) &\ni L^{-1}\nabla f(x)
                \\
                L(x^+ - x) + \partial g(x^+) &\ni - \nabla f(x)
                \\
                L(x - x^+) &\in \nabla f(x) + \partial g(x^+)
                \\
                \mathcal G_L(x) &\in \nabla f(x) + \partial g(x^+). 
            \end{align*}
        \end{proof}

        \begin{lemma}[Linearized Gradient Mapping Lower Bound]
            Continue from 
            \hyperref[lemma:grad_map_lemma_first]{definition \ref*{lemma:grad_map_lemma_first}}, 
            with $x^+ = \mathcal T(x)$, the gradient mapping satisfies the inequality for all $z$: 
            \begin{align*}
                h(z) &\ge
                h(z) - \frac{L}{2}\Vert x - x^+\Vert^2
                \ge h(x^+) + \langle \mathcal G_L(x), z - x\rangle. 
            \end{align*}
        \end{lemma}
        \begin{proof}
            Directly from the $L$-smoothness of $f$, convexity of $g, f$, we have the list of inequalities: 
            \begin{align*}
                &f(x^+) \le 
                f(x) + \langle \nabla f(x), x^+ - x\rangle
                + \frac{L}{2}\Vert x - x^+\Vert^2, 
                \\
                &f(x) + \langle \nabla f(x), z - x\rangle 
                \le f(z), 
                \\
                &g(x^+) \le 
                g(z) + \langle \partial g(x^+), x^+ - z\rangle. 
            \end{align*}
        \end{proof}

    \subsection{Varieties of Nesterov accelerated gradient}\label{sec:AG_varieties}
        Here, the acrynonym: ``AG" stands for accelerated gradient. 
        \subsubsection{AG Abstract Forms}
            In this section, we list different varieties of the Nesterov accelerated method. 
            \begin{definition}[Nestrov 2.2.7]\label{def:Nes2.2.7}
                Let $f$ be a $L$ Lipschitz smooth and $\mu\ge 0$ strongly convex function. 
                Choose $x_0$, $\gamma_0 > 0$, set $v_0 = x_0$, for iteration $k\ge 0$, it
                \begin{enumerate}
                    \item[1.] computes $\alpha_k \in (0, 1)$ by solving $L\alpha_k^2 = (1 - \alpha_k)\gamma_k + \alpha_k \mu$; 
                    \item[2.] sets $\gamma_{k + 1} = (1 - \alpha_k)\gamma_k + \alpha_k \mu$;
                    \item[3.] chooses $y_k = (\gamma_k + \alpha_k \mu)(\alpha_k \gamma_k v_k + \gamma_{k + 1}x_k)$. Compute $f(y_k)$ and $\nabla f(y_k)$; 
                    \item[4.] finds $x_{k + 1}$ such that $f(x_{k + 1}) \le f(y_k) - (2L)^{-1} \Vert \nabla f(y_k)\Vert^2$; 
                    \item[5.] sets $v_{k + 1} = \gamma_{k+1}^{-1}((1 - \alpha_k)\gamma_kv_k + \alpha_k \mu y_k - \alpha_k \nabla f(y_k))$. 
                \end{enumerate}
            \end{definition}
            \begin{remark}
                This is in Nesterov's book \cite[(2.2.7)]{nesterov_lectures_2018}. 
                For more context, the sequence $\alpha_k \in (0, 1)$ such that $\sum_{k = 1}^{\infty}\alpha_k = \infty$, and recursively we have $\lambda_{k + 1} = (1 - \alpha_k)\lambda_k$. 
                the sequence $\lambda_k$ is called an Nesterov estimating sequence. 
                It is the most generic algorithm in his book about accelerated gradient method. 
                The genericity of the algorithm is provided by item 4., which is the a special case of the smooth descent lemma.

@@

            \begin{definition}[AG Proximal Gradient PPM Generic Form]

            \end{definition}

@@

        \subsubsection*{Useful Observations about these Generic forms}\label{sec:AG_useful_information}

@@

                $L$-Lipschitz smooth, 

@@

                Similarly, when $\mu = 0$, we have from the first step that 
                    v_{k + 1} 
                    &= ((1 - \alpha_k)\gamma_k)^{-1}
                        (1 - \alpha_k)\gamma_k v_k
                        + \alpha_k \mu y_k - \alpha_k \nabla f(y_k)
                    v_k - \alpha_k((1 - \alpha_k)\gamma_k)^{-1}\nabla f(y_k)
                which is the same as the AG generic PPM form where 

@@

            the AG generic performs PPM on $\phi_t$ to produce $x_{t + 1}$ so the PPM Lyapunov inequality from 
            \hyperref[thm:ppm_descent_ineq]{theorem \ref*{thm:ppm_descent_ineq}} applies,

@@

@@

                - \frac{1}{2}\Vert x_{t} - x_*\Vert^2  =: \Upsilon_{1, t + 1}^{\text{AG}}

@@

        \begin{theorem}[Generic AG Convergence]\label{thm:generic_smooth_ag_convergence}
            If there is a choice of $\eta_i, \tilde \eta_i$ in \hyperref[def:agg_ppm]{definition \ref*{def:agg_ppm}} such that we have 
            \begin{align*}
                \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)
                \Upsilon_{2, t + 1}^{\text{AG}} + 
                \Upsilon_{1, t + 1}^{\text{AG}} \le 0, 
            \end{align*}
            then with 
            \begin{align*}
                \Phi_t &= \left(
                    \sum_{i = 1}^{t} \tilde\eta_{i}
                \right) (f(z_t) - f(x_*)) + \frac{1}{2}\Vert x_t - x_*\Vert^2 \quad \forall t \in \NN
                \\
                \Phi_0 &= \frac{1}{2}\Vert x_t - x_*\Vert^2, 
            \end{align*}
            we have $\Phi_{t + 1} - \Phi_t \le 0$, which allows for a convergence rate of $\mathcal O \left(\sum_{i = 1}^{T} \eta_i^{-1}\right)$ for $f(z_T) - f(x_*)$. 
        \end{theorem}
        \begin{proof}
            By definition we have
            {\footnotesize
            \begin{align*}
                \Phi_{t + 1} - \Phi_t 
                &= 
                \left(
                    \sum_{i = 1}^{t+1} \tilde\eta_{i}
                \right) (f(z_{t + 1}) - f(x_*)) 
                - 
                \left(
                    \sum_{i = 1}^{t} \tilde\eta_{i}
                \right) (f(z_{t}) - f(x_*)) 
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                - \frac{1}{2}\Vert x_{t + 1} - x_*\Vert^2
                \\
                &= 
                \tilde \eta_{t + 1} (f(z_{t + 1}) - f(z_*))
                +
                \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)(f(z_{t + 1}) - f(z_t))
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                - \frac{1}{2}\Vert x_{t + 1} - x_*\Vert^2
                \\
                &= \left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right)\Upsilon_{2, t + 1}^{\text{AG}} + \Upsilon_{1, t + 1}^{\text{AG}} \le 0. 
            \end{align*}
            }
            The derivation for the convergence raet is direct and similar to 
            \hyperref[thm:ppm_convergence_rate]{theorem \ref*{thm:ppm_convergence_rate}} 
            Therefore, if we can identify parameter for the generic algorithm that asserts the condition above, then we have convergence for the algorithm. 
        \end{proof}

    \subsection{Specific Varieties of Accelerated Gradient Method}
        In this section we state some specific variant of the Nesterov accelerated gradient method. 
        The "AG" stands for accelerated gradient.
